---
title: "Mono Lake Water Levels"
description: |
  A visual history
author:
  - name: Allison Gracer
date:
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Attach packages
library(tidyverse)
library(janitor)
library(here)
library(broom)
```



```{r}
# read in the data
mono_lake <- read_csv(here("data", "mono_lake_annual_levels.csv"),
                           skip = 5) %>% 
  clean_names()
```
 
 
 
 
```{r fig.align = "center"}
# visualize the data
ggplot(data = mono_lake, aes(x = year, y = lake_level_feet_above_sea_level)) + # create ggplot
  geom_line(color = "cadetblue4") + # specify plot type, and customize line color
  geom_ribbon(aes(ymin = 6350, # create fill for the water levels, set min
                  ymax = lake_level_feet_above_sea_level), #set fill max
              fill = "lightcyan2") + # customize fill color
  labs(x= "\nYear\n", # label x-axis
       y = "\nLake level\n(Feet above sea level)\n", # label y-axis
       title = "Mono Lake Levels 1850-2017") + # create a title
  theme_bw() + # customize theme
  theme(plot.title = element_text(hjust = 0.5, # center the tile
                                  size = 14, # change title font size
                                  face = "bold")) + # change title font to bold
  theme(plot.background = element_rect(fill = "burlywood"), # change plot background color
        panel.background = element_rect(fill = "ivory1")) + #change panel background color
  theme(plot.margin = margin(t = 10, # top margin
                             r = 30, # right margin
                             b = 5, # bottom margin
                             l = 1)) + # left margin
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_x_continuous(limit = c(1850,2017), # set limits of x-axis
                     breaks = c(1850, 1875, 1900, 1925, 1950, 1975, 2000, 2017), # specify breaks
                     expand = c(0,0)) + # adjust axis margins
  scale_y_continuous(limit = c(6350, 6435), # set limits of y-axis
                     breaks = c(6350, 6370, 6390, 6410, 6430),  # specify breaks
                     expand = c(0,0)) + # adjust axis margins
  geom_line(aes(y = 6377), # add another line to the graph to indicate important water levels
            color = "sienna4", # customize line color
            linetype = "dotted") + # customize line type
  geom_line(aes(y = 6360), # add another line to the graph to indicate important water levels
            color = "sienna4", # customize line color
            linetype = "dotted") + # customize line type
  annotate("text", # add text labels, indicate location, color, and size
           label = "6,377 feet - land bridges emerge",
           x = 1875, # label position on x-axis
           y = 6380, # label position on y-axis
           color ="sienna4", #label font color
           size = 2.75) + # label font size
  annotate("text", # add text labels, indicate location, color, and size
           label = "6360 feet - birds food sources decline", # label title
           x = 1879, # label position on x-axis
           y = 6363, # label position on y-axis
           color = "sienna4", # label font color
           size = 2.75) + # label font size
  geom_segment(aes(x = 1941, # add an arrow to indicate 1941 water diversion to LA
                   xend = 1941,
                   y = 6420,
                   yend = 6429),
               color = "sienna4",
               arrow = arrow(ends = "first", # set the arrow facing down
                             type = "open", # set arrow point to open (not filled in)
                             length = unit(.3, "cm"))) + # set arrow point length
  annotate("text", # add text labels, indicate location, color, and size
           label = "1941 - water diversion to LA begins",
           x = 1941, # label position on x-axis
           y = 6432, # label position on y-axis
           color = "sienna4", # label font color
           size = 2.75) + # label font size
  geom_segment(aes(x = 1983, # add an arrow to indicate 1983 CA Supreme Court ruling
                   xend = 1983, # x-axis end point
                   y = 6381, # y-axis start point
                   yend = 6390), # y-axis line end point
               color = "sienna4", # arrow color
               arrow = arrow(ends = "first", # set the arrow facing down
                             type = "open", # set arrow point to open (not filled in)
                             length = unit(.3, "cm"))) +  # set arrow point length
    annotate("text", # add text labels, indicate location, color, and size
           label = "1983 - CA Supreme Court protection",
           x = 1990, # label position on x-axis
           y = 6393, # label position on y-axis
           color = "sienna4", # label font color
           size = 2.75)

```
 
**Figure 1:** This figure depicts the recent history of Mono Lake water levels from 1850-2017. In 1941 the Los Angeles Department of Water & Power (DWP) began diverting water to supply the growing demand of water for Los Angeles. This diversion of water had profound effects on Mono Lake water and salinity levels, and resulted in a number of ecological impacts. Around 1976 the lake levels dropped to 6,377 feet, land bridges emerged allowing predators access to islands that had previously been safe nesting grounds for California Gulls. In 1983 the California Supreme Court sided with the Mono Lake Committee (est. 1979) ruling that the state of California had an obligation to protect Mono Lake, even if that meant finding other sources of water for Los Angeles. A study published in 1987 by the Mono Basin Ecosystem Study Committee determined a decrease in bird food sources would occur at lake levels of 6,360 feet. 

**Data Citation:** Mono Basin Clearinghouse








